---
title: "Borders Cancer Rates Analysis"
output:
  html_document:
  html_notebook: default

---

```{r, include=FALSE}
library(tidyverse)
library(janitor)
library(babynames)
library(tidytext)
```


```{r, include=FALSE}
#load in data and alter variable names
geo_codes <- read_csv("geography_codes.csv") %>% clean_names()

hb <- read_csv("health_boards.csv") %>%  clean_names()

scotland <- read_csv("scotland_wide.csv") %>%  clean_names()

regions <- read_csv("cancer_network_region.csv") %>% clean_names()


```

```{r, include=FALSE}
 #find and store the value (as char) for NHS borders health board
hb_code_borders <- geo_codes %>% 
  filter(hb_name == "NHS Borders") %>% 
  pull(hb)
```


```{r, include=FALSE}
#Account the number of cancer cases (male & female) per year

hb_graph <- hb %>% 
  filter(hb == hb_code_borders) %>% 
  filter(sex != "All") %>% 
  group_by(year, sex) %>% 
  summarise(count = sum(incidences_all_ages))
```

```{r, echo=FALSE}
  #plot this as a line graph
hb_graph %>% 
  ggplot(aes(x = year, y = count, colour = sex)) +
  geom_line() +
  scale_colour_manual(values=c("orange","lightblue")) +
  geom_point(size=2, shape=21, fill = "white") +
  labs(
        y = "Cancer Cases",
        x = "Year",
        title = "Cancer Rates",
        subtitle = "By Gender 1994-2018"
      ) +
  theme(plot.title=element_text(hjust = 0.5),
        plot.subtitle=element_text(hjust = 0.5)) +
   scale_x_continuous(breaks = seq(min(hb$year),max(hb$year))) +
    scale_y_continuous(breaks = seq(0,1500 , by = 100)) +
  theme(axis.text.x = element_text( angle = 45, hjust = 1 )) +
  theme(legend.title = element_blank())


  
```

&nbsp;

Clearly the cancer rates have increased between 1994-2018.
The rate amongst females has increased by 41%, more alarmingly the rate amongst males has increased by 128%

&nbsp;




```{r, include=FALSE}
#find the 10 busiest cancer centres (1994-2018) split by gender:
#store the top ten sites for females
female_count <- hb %>% 
  filter(hb == hb_code_borders) %>%
  filter(sex != "All") %>% 
  filter(cancer_site != "All cancer types") %>%
  group_by(cancer_site, sex) %>% 
  summarise(count = sum(incidences_all_ages)) %>% 
  arrange(desc(count)) %>%
  filter(sex == "Female") %>% 
  head(10)

#store the top ten sites for males
male_count <- hb %>% 
  filter(hb == hb_code_borders) %>%
  filter(sex != "All") %>% 
  filter(cancer_site != "All cancer types") %>%
  group_by(cancer_site, sex) %>% 
  summarise(count = sum(incidences_all_ages)) %>% 
  arrange(desc(count)) %>%
  filter(sex == "Male") %>% 
  head(10)

combined_mf <- bind_rows(female_count, male_count)

```

```{r, echo=FALSE}
  #plot this as a bar graph
combined_mf %>% 
  mutate(cancer_site = reorder_within(cancer_site, count, sex)) %>% 
  ggplot(aes(x = reorder(cancer_site, -count), y = count)) +
  geom_col(aes(fill = sex), col = "black") +
  scale_x_reordered() +
  facet_wrap(~sex, scales = "free_x") +
   geom_text(aes(label = count),
            size = 2,
            position = position_stack(vjust = 0.5),
            colour = "black") +
  theme(axis.text.x = element_text( angle = 45, hjust = 1, size = 6.5 )) +
   scale_fill_manual(values=c("orange","lightblue")) +
 theme(legend.title = element_blank()) +
  labs(
        y = "Cancer Cases",
        x = "Type",
        title = "Cancer Rates by Type",
        subtitle = "Most Frequent Types 
        Separated by Gender 1994-2018"
      ) +
  theme(plot.title=element_text(hjust = 0.5),
        plot.subtitle=element_text(hjust = 0.5))

  
  
  
  
#   ggplot(aes(x = reorder(cancer_site, -count), y = count, 
#              fill = sex)) +
#   geom_col(col = "black") +
#    geom_text(aes(label = count),
#             size = 2,
#             position = position_stack(vjust = 0.5),
#             colour = "black") +
# theme(axis.text.x = element_text( angle = 45,  hjust = 1, size = 6.5 )) +
#    scale_fill_manual(values=c("orange","lightblue")) +
#  theme(legend.title = element_blank()) +
#   labs(
#         y = "Cancer Cases",
#         x = "Type",
#         title = "Cancer Rates by Type",
#         subtitle = "Most Frequent Types Separated by Gender 1994-2018"
#       ) +
#   theme(plot.title=element_text(hjust = 0.5),
#         plot.subtitle=element_text(hjust = 0.5))
# 
#   
  
```

&nbsp;

The above gives the busiest cancer sites across 1994-2018, it attempts to show where resources should be prioritised.  Non-melanoma skin cancer is the most prevalent cancer type in the Borders.

&nbsp;



```{r, include=FALSE}
#find the 10 busiest cancer centres for 2018 (latest year of data):
female_count_2018 <- hb %>% 
  filter(hb == hb_code_borders) %>%
  filter(sex != "All") %>% 
    filter(cancer_site != "All cancer types") %>%
   filter(year == max(year)) %>% 
  group_by(cancer_site, sex) %>% 
  summarise(count = sum(incidences_all_ages)) %>% 
  arrange(desc(count)) %>%
  filter(sex == "Female") %>% 
  head(10)

(male_count_2018 <- hb %>% 
  filter(hb == hb_code_borders) %>%
  filter(sex != "All") %>%
  filter(cancer_site != "All cancer types") %>%
  filter(year == max(year)) %>% 
  group_by(cancer_site, sex) %>% 
  summarise(count = sum(incidences_all_ages)) %>% 
  arrange(desc(count)) %>%
  filter(sex == "Male") %>% 
  head(10))

combined_mf_2018 <- bind_rows(female_count_2018, 
                              male_count_2018)



```

```{r, echo=FALSE}
  #plot this as a bar graph
combined_mf_2018 %>% 
    mutate(cancer_site = reorder_within
           (cancer_site, count, sex)) %>% 
  ggplot(aes(x = reorder(cancer_site, -count), y = count)) +
  geom_col(aes(fill = sex), col = "black") +
  scale_x_reordered() +
  facet_wrap(~sex, scales = "free_x") +
     geom_text(aes(label = count),
            size = 2,
            position = position_stack(vjust = 0.5),
            colour = "black") +
  theme(axis.text.x = element_text( angle = 45, 
                                    hjust = 1,  size = 6.5 )) +
    scale_fill_manual(values=c("orange","lightblue")) +
 theme(legend.title = element_blank()) +
  labs(
        y = "Cancer Cases",
        x = " Type",
        title = "Cancer Rates by Type",
        subtitle = "Most Frequent Types Separated by Gender 2018"
      ) +
  theme(plot.title=element_text(hjust = 0.5),
        plot.subtitle=element_text(hjust = 0.5))


  
#   ggplot(aes(x = reorder(cancer_site, -count), y = count, 
#              fill = sex)) +
#   geom_col(col = "black") +
#    geom_text(aes(label = count),
#             size = 2,
#             position = position_stack(vjust = 0.5),
#             colour = "black") +
# theme(axis.text.x = element_text( angle = 45,  hjust = 1, size = 6.5)) +
#    scale_fill_manual(values=c("orange","lightblue")) +
#  theme(legend.title = element_blank()) +
#   labs(
#         y = "Cancer Cases",
#         x = " Type",
#         title = "Cancer Rates by Type",
#         subtitle = "Most Frequent Types Separated by Gender 2018"
#       ) +
#   theme(plot.title=element_text(hjust = 0.5),
#         plot.subtitle=element_text(hjust = 0.5))

```

&nbsp;

The above gives the busiest cancer sites in 2018, it may be a better indication as to the current needs for cancer patients.  Again non-melanoma skin cancer was the most prevalent cancer type in the Borders in 2018.

&nbsp;




```{r, include=FALSE}
#find the number of cases at each cancer site in the Borders
hb_borders_agg_type <- hb %>% 
  filter(hb == hb_code_borders) %>%
  filter(sex == "All") %>% 
  group_by(cancer_site) %>% 
  summarise(count = sum(incidences_all_ages))

#find the number of cases  each year in the Borders
hb_borders_agg_year <- hb %>% 
  filter(hb == hb_code_borders) %>%
  filter(sex == "All") %>% 
  group_by(year) %>% 
  summarise(count = sum(incidences_all_ages))

```

```{r, include=FALSE}
#find the number of cases at each cancer site in Scotland
scotland_agg_type <- scotland %>% 
  filter(sex == "All") %>% 
  group_by(cancer_site) %>% 
  summarise(count = sum(incidences_all_ages))

#find the number of cases at each year in Scotland
scotland_agg_year <- scotland %>% 
  filter(sex == "All") %>% 
  group_by(year) %>% 
  summarise(count = sum(incidences_all_ages))
```

```{r, include=FALSE}


# combine the count per year data for the Borders and Scotland
years_combined <- inner_join(hb_borders_agg_year, scotland_agg_year, by = "year")


years_combined <- years_combined %>% 
  #rename default named columns
  rename(borders = count.x) %>% 
  rename(scotland_wide = count.y) %>% 
  #calculate total cases
  mutate(total_cases = borders + scotland_wide)


#make the data tidy to allow easier plotting: create a single column for location and place cancer case numbers in an appropriately named column 
years_combined <- years_combined %>% 
  pivot_longer(
    cols = borders:scotland_wide,
    names_to = "location",
    values_to = "cancer_cases"
  )

#find the case number as a percentage of the total cases
years_combined <- years_combined %>% 
  mutate(percentage_cases =
  round(((cancer_cases/total_cases)*100), digits = 2))


  
```

```{r, echo=FALSE}
#plot the cancer rates in the Borders compared to those Scotland wide
  years_combined %>% 
  ggplot(aes(x = year, y = cancer_cases, fill = location)) +
  geom_col(col = "black", width = 1) +
  geom_text(aes(label = paste(percentage_cases,"%")),
            size = 2.5,
            position = position_stack(vjust = 0.5),
            hjust = -0.5,
            colour = "black", angle = 90)  +
  ylim(0, max(years_combined$cancer_cases) * 1.2) +
    labs(
        y = "Cancer Cases",
        x = "Year",
        title = "Cancer Rates by Location",
        subtitle = "Presented as a proportion 1994-2018"
      ) +
  scale_fill_manual(values=c("orange","lightblue")) +
  theme(plot.title=element_text(hjust = 0.5),
        plot.subtitle=element_text(hjust = 0.5)) +
   scale_x_continuous(breaks = seq(min(years_combined$year),max(years_combined$year))) +
    #scale_y_continuous(breaks = seq(0,100000 , by = 10000)) +
  theme(axis.text.x = element_text( angle = 45, hjust = 1 )) +
  theme(legend.title = element_blank())


```

&nbsp;

The proportion of cancer cases between 1994-2018 in the Borders has remained between 2.05-2.61 percent 

&nbsp;



```{r, include=FALSE}
#find the number of cases at each cancer site in the region "South East of Scotland"
regions_agg_type <- regions %>% 
  filter(region == "South East of Scotland") %>%
  filter(sex == "All") %>% 
  group_by(cancer_site) %>% 
  summarise(count = sum(incidences_all_ages))


#find the number of cases  each year in the region "South East of Scotland"
regions_agg_year <- regions %>% 
  filter(region == "South East of Scotland") %>%
  filter(sex == "All") %>% 
  group_by(year) %>% 
  summarise(count = sum(incidences_all_ages))


```

```{r, include=FALSE}
# combine the count per year data for the Borders and the region in which it belongs
years_combined_region <- inner_join(hb_borders_agg_year, 
                                    regions_agg_year, by = "year")


years_combined_region <- years_combined_region %>% 
  #rename default named columns
  rename(borders = count.x) %>% 
  rename(south_scotland = count.y) %>% 
  #calculate total cases
  mutate(total_cases = borders + south_scotland)

#make the data tidy to allow easier plotting: create a single column for location and place cancer case numbers in an appropriately named column 
years_combined_region <- years_combined_region %>% 
  pivot_longer(
    cols = borders:south_scotland,
    names_to = "location",
    values_to = "cancer_cases"
  )


#find the case number as a percentage of the total cases
years_combined_region <- years_combined_region %>% 
  mutate(percentage_cases =
  round(((cancer_cases/total_cases)*100), digits = 2))


```

```{r, echo=FALSE}
#plot the cancer rates in the Borders compared to those in the south of Scotland 
  years_combined_region %>% 
  ggplot(aes(x = year, y = cancer_cases, fill = location)) +
  geom_col(col = "black", width = 1) +
  geom_text(aes(label = paste(percentage_cases,"%")),
            size = 2.5,
            position = position_stack(vjust = 0.5),
            hjust = -0.5,
            colour = "black", angle = 90)  +
  ylim(0, max(years_combined_region$cancer_cases) * 1.2) +
    labs(
        y = "Cancer Cases",
        x = "Year",
        title = "Cancer Rates by Location (Region - S.E. Scotland)",
        subtitle = "Presented as a proportion 1994-2018"
      ) +
  scale_fill_manual(values=c("orange","lightblue")) +
  theme(plot.title=element_text(hjust = 0.5),
        plot.subtitle=element_text(hjust = 0.5)) +
   scale_x_continuous(breaks = seq(min(years_combined$year),max(years_combined$year))) +
    #scale_y_continuous(breaks = seq(0,100000 , by = 10000)) +
  theme(axis.text.x = element_text( angle = 45, hjust = 1 )) +
  theme(legend.title = element_blank())
```

&nbsp;

The proportion of cancer cases between 1994-2018 in the Borders compared to those in the region South East Scotland has remained between 7.14-8.94 percent 



```{r}
# combined_mf %>%
#   mutate(cancer_site = reorder_within(cancer_site, count, sex)) %>% 
#   ggplot(aes(x = reorder(cancer_site, -count), y = count)) +
#   geom_col() +
#   scale_x_reordered() +
#   facet_wrap(~sex, scales = "free_x") +
#   theme(axis.text.x = element_text( angle = 45, hjust = 1 ))
```

